The hedging hyperprior for the exponential distribution

### 2023-03-06 text/23/gamma-hedging-hyperprior.Rmd
Added contour plot to gamma-hedging-hyperprior

```{r setup}
suppressMessages(suppressWarnings(library(tidyverse)))
knitr::opts_chunk$set(echo=TRUE, fig.width=8, fig.height=2)
```

@@

```{r}

a0 <- 10
b0 <- 30

n <- 30
xbar <- 0.7

@@

x_grid <- seq(0.01, 8, by=0.01)

@@

prior05 <- 1/qgamma(c(0.95), a0, b0) 
prior95 <- 1/qgamma(c(0.05), a0, b0)

post05 <- 1/qgamma(c(0.95), a1, b1) 
post95 <- 1/qgamma(c(0.05), a1, b1)

@@

```{r}
f_grid <- f(xbar, x_grid)^10
x_color <- "white"
data.frame(x=x_grid, y=f_grid) %>%
  ggplot(aes(x, y)) +
    geom_line() + xlab(" ") + ylab(" ")
```

@@

### The power prior

The power prior was developed to combat the excessive influence of an informative prior when the data conflicts with that prior.

Instead of the standard Bayesian approach where you multiply the Likelihood (L) by the prior (g), you use

$h(\theta|\mathbf{X}) \propto L(\mathbf{X}|\theta)g(\theta)^\tau g_0(theta)$

where $g_0$ is a non-informative prior and $\tau$ is between 0 and 1.

For $\tau$=1, the prior is used at full strength. When $\tau$=0, the informative prior effectively disappears, leaving you with just the non-informative prior. For values of $\tau$ between 0 and 1, the informative prior is weakened, but still exerts some pull on the posterior distribution.

For intermediate values of $\tau$, the formulation is not consistent with Bayes rule and you need to include a "fudge factor" of

$\int g(\theta)^{\tau}g_0(\theta)d\theta$

A simpler approach is to include $\tau$ in the prior distribution and then place a hyperprior on $\tau$.

The parameter $\alpha$ in an inverse gamma distribution can be thought of as a prior sample size. Large values of $\alpha$ produce informative priors that are equivalent to adding $\alpha$ pseudo-data values. If you multiply $\alpha$ by $\tau$, you can reduce the amount of pseudo data being added.

@@

>>>>

```{r setup}
suppressMessages(suppressWarnings(library(MASS)))
suppressMessages(suppressWarnings(library(tidyverse)))
knitr::opts_chunk$set(echo=TRUE, fig.width=8, fig.height=2)
```

@@

```{r}
x_max <- 3

a0 <- 41
b0 <- 44

n <- 60
xbar <- 2.1

@@

x_grid <- seq(0.01, x_max, length=1000)

@@

prior05 <- 1/qgamma(c(0.975), a0, b0) 
prior95 <- 1/qgamma(c(0.025), a0, b0)

post05 <- 1/qgamma(c(0.975), a1, b1) 
post95 <- 1/qgamma(c(0.025), a1, b1)

@@

```{r}
f_grid <- f(xbar, x_grid)^n
x_color <- "white"
data.frame(x=x_grid, y=f_grid) %>%
  ggplot(aes(x, y)) +
    geom_line() + xlab(" ") + ylab(" ")
```

@@

@@

When the likelihood of your data and your informative prior are in conflict, you may wish to downweight the informative prior. You can do this by including a hedging hyperprior.

Define your prior distribution as an inverse gamma distributions with parameters $1+\lambda (\alpha-1)$ and $\lambda \beta$ where $\lambda$ ranges between 0 and 1. Then place a hyperprior on the parameter $\lambda$. The uniform(0,1) distribution is a reasonable choice for the hyperprior, but others are possible.

```{r}
knitr::opts_chunk$set(echo=TRUE, fig.width=8, fig.height=8)
```

```{r}
tau_grid <- seq(0, 1, length=60)
theta_grid <- seq(0, x_max, length=60)

gstar <- function(theta, tau) {g(theta, tau*(a0-1)+1, tau*(b0-1)+1)}
density <- outer(theta_grid, tau_grid, gstar)
op <- par(bg = "white")
z_max <- max(density, na.rm=TRUE)
persp(
  theta_grid, 
  tau_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="theta",
  ylab="tau",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

It is hard to see at this scale, but the prior distribution is higher for small values of $\tau$ when $\theta4 is very large or very small. This becomes more obvious if you remove the middle of the surface and rescale the Z-axis.

```{r}
density <- outer(theta_grid, tau_grid, gstar)
density[11:40, ] <- NA
op <- par(bg = "white")
z_max <- max(density, na.rm=TRUE)
persp(
  theta_grid, 
  tau_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="theta",
  ylab="tau",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

The likelihood function is constant across all values of $\tau$, so its surface looks like the following.

```{r}
tau_grid <- seq(0, 1, length=60)
theta_grid <- seq(0, x_max, length=60)

fstar <- function(theta, tau) {f(xbar, theta)^n}
likelihood <- outer(theta_grid, tau_grid, fstar)
z_max <- max(likelihood, na.rm=TRUE)
persp(
  theta_grid, 
  tau_grid, 
  likelihood, 
  zlim=c(0, 1.1*z_max),
  xlab="theta",
  ylab="tau",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

The product of the prior and the likelihood is proportional to the posterior distribution.

```{r}
tau_grid <- seq(0, 1, length=60)
theta_grid <- seq(0, x_max, length=60)

hstar <- function(theta, tau) {g(theta, tau*(a0-1)+1, tau*(b0-1)+1)*f(xbar, theta)^n}
density <- outer(theta_grid, tau_grid, hstar)
z_max <- max(density, na.rm=TRUE)
persp(
  theta_grid, 
  tau_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="theta",
  ylab="tau",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

A contour plot shows where the probability is concentrated. The labels on the contour lines represent how much probabilty is found inside the contour line.

```{r}
density <- density / sum(density, na.rm=TRUE)
density %>%
  as.vector %>%
  sort %>%
  rev -> sorted_density
cumulative_p <- 0
for (p in sorted_density) {
  cumulative_p <- p + cumulative_p
  density[density==p] <- cumulative_p
}
contour(
  theta_grid, 
  tau_grid, 
  density,
  labcex=1,
  levels=c(10, 25, 50, 75, 90)/100,
  xlab="theta",
  ylab="tau")
```

You can see that the hedging hyperprior did its job. The most probable values of $\tau$ are small and the prior mean (`r b0/(a0-1)`) is largely ignored in favor of the mean of the data (`r xbar`).
